Gamma-Ray Burst Afterglows In The Multi-Messenger Era: Numerical Models and Closure Relations

### 2019-09-23 structuredJets.tex
Almost there...

Interpreting the afterglow as arising from a structured jet allows one to quickly draw the following conclusions:
\begin{enumerate}
	\item The spectral slope $\beta = 0.585 \pm 0.005$ throughout the entire evolution, identifying the relevant spectral regime as G: $\nu_m < \nuobs < \nu_c$, and $p = 2.17 \pm 0.01$ \citep{Troja:2019ab}.
	\item The initial rising ($\alpha > 0$) light curve with declining spectrum ($\beta < 0$) immediately identify this as a \emph{misaligned} evolution ($\thobs > \thC$).  Aligned jets may only have a rising light curve when viewed at frequencies below the peak frequency, that is when $\beta < 0$.
	\item The rising light curve maintains $\alpha  = 0.90 \pm 0.06$ \citep{Troja:2019ab}.  Using the closure relations from Table \ref{tab:slopes} we can identify this as the \emph{structured} phase and determine $\geff = 8.2 \pm 0.5$.  If the light curve underwent an \emph{far-off-axis} phase, it concluded before the afterglow was detected: $\tW < 9$d.  The peak can be identified as the jet-break: $\tb = 164 \pm 12$ d.
	\item In the context of a Gaussian jet, the measurement of $\geff$ leads immediately to $\thobs / \thC = 5.7 \pm 0.2$ using Equation \eqref{eq:geff}.  A $b=2$ power law jet cannot achieve this $\geff$ for any reasonable value of $\thobs / \thC$.  A $b=6$ power law jet requires $\thobs/\thC = 14 \pm 1$.
\end{enumerate}
To go further we first need to assume a value for $E_0 / n_0$. For demonstration we use the \citet{Fong:2015aa} fiducial values of $E_0 \sim 2\times 10^{51}$erg and $n_0 \sim 10^{-2}$cm$^{-3}$, assuming 1 dex uncertainty on both parameters.
\begin{enumerate}
	\setcounter{enumi}{4}
	\item We can use the jet break $\tb$ measurement to determine $\thobs + \thC$ using Equation \eqref{eq:tbout}. Using our assumed values for  $E_0$ and $n_0$ leads to $\thobs + \thC \approx 0.92 \pm 0.38$ rad ($53^\circ \pm 22^\circ$). This does not depend on the jet model.
	\item For a Gaussian jet the combined measurements of $\thobs +\thC$ and $\thobs/\thC$ then imply $\thC \approx 0.14 \pm 0.06$ rad ($8.0^\circ  \pm 3.4^\circ$) and $\thobs \approx 0.78 \pm 0.32$ rad ($45^\circ \pm 18^\circ$).
	\item For a $b=6$ power law jet we get $\thC \approx 0.063 \pm 0.026$ rad ($3.6^\circ  \pm 1.5^\circ$) and $\thobs \approx 0.86 \pm 0.35$ rad ($49^\circ \pm 20^\circ$).
\end{enumerate}

@@

%TODO FINISH THIS PARAGRAPH
Analysis of the \gwbns{} gravitational wave signal produced estimates of the inclination angle $\iota$ between the binary neutron stars' orbital plane and our line of sight.

This is in agreement with the LIGO measurement $\thobs \approx 28^\circ \pm 15^\circ$ (assuming the SHoES $H_0$).

For a more detailed analysis we performed Bayesian parameter estimation utilizing \afterglowpy{} and the \emcee{} Markov chain Monte Carlo sampler \citep{Foreman-Mackey:2013aa}.  We used the light curve data reported in \citet{Troja:2019ab} with additional \hubble{} observations reported in \citet{Lamb:2019aa}. Reported measurement uncertainties were assumed to be independent and Gaussian, upper limits were treated as observations of zero flux with a 1-sigma Gaussian uncertainty.

We fit both the Gaussian and power law structured jet models, fixing $\xi_N = 1$ and $\dL = 1.23 \times 10^{26}$cm.  The fit parameters, their priors, and their marginalized posteriors are given in Table \ref{tab:MCMC}. Both models were run with 300 walkers for 64 000 iterations, discarding the first 16 000 iterations as a burn-in.  The viewing angle prior is the posterior found in analysis of the \gwbns{} gravitational wave signal assuming the value of $H_0$ reported by the \planck{} collaboration \citep{Abbott:2017aa, Planck-Collaboration:2016aa, Troja:2018aa}.

\begin{deluxetable*}{CCCCCC}
	\tablecaption{Parameter estimation priors and marginalized posteriors for the \gwbns{} afterglow using the \afterglowpy{} Gaussian and power law jet models, including viewing angle constraints from LIGO assuming the \planck{} value of $H_0$.  Given posterior values for each model are the median, 16\%, and 84\% quantiles.  Parameters in the lower section are derived from the posterior distributions of the fit parameters in the upper sections. \label{tab:MCMC}}
	\tablehead{\colhead{Parameter}& \colhead{Unit} & \colhead{Prior Form} & \colhead{Bounds} & \colhead{Gaussian Jet Posterior} & \colhead{Power Law Jet Posterior}}
	\startdata
		\thobs	& \text{rad	} & \sin \thobs \times p_\mathrm{LIGO}(\cos \thobs)	& [0, 0.8] 	& 0.48^{+0.08}_{-0.09} & 0.36^{+0.14}_{-0.11} \\
		\log_{10} E_0	& \text{erg	}	& \text{uniform}	& [45, 57] 			& 52.92^{+0.62}_{-0.49} & 53.01^{+0.99}_{-0.70}\\
		\thC			& \text{rad	}	& \text{uniform}	& [0.01, $\pi/2$]		& 0.07^{+0.01}_{-0.01} & 0.10^{+0.06}_{-0.03}\\
		\thW			& \text{rad}	& \text{uniform}	& [0.01, 12$\thC$]	& 0.77^{+0.43}_{-0.38} & 0.18^{+0.07}_{-0.05}\\
		b			& - 			& \text{uniform}	& [0, 10] 			& - & 7.15^{+2.02}_{-1.12}\\
		\log_{10}n_0	& \text{cm}^{-3} & \text{uniform}	& [-10, 10] 		& -1.94^{+0.62}_{-0.65} & -2.66^{+1.03}_{-1.11}\\
		p			& - 			& \text{uniform}	& [2, 5] 			& 2.17^{+0.01}_{-0.01} & 2.17^{+0.01}_{-0.01}\\
		\log_{10}\epse& - 			& \text{uniform}	& [-5, 0] 			& -1.24^{+0.48}_{-0.69} &-1.30^{+0.66}_{-1.07} \\
		\log_{10}\epsB	& - 			& \text{uniform}	& [-5, 0] 			& -4.32^{+0.72}_{-0.49} & -3.84^{+1.01}_{-0.80}\\
		\hline
		\log_{10}\Etot	& \text{erg}	&	-		&	-			& 50.61^{+0.58}_{-0.44} &50.85^{+0.93}_{-0.63} \\
		\thobs/\thC	& -			&	-		&	-			& 6.76^{+0.21}_{-0.21} & 3.78^{+0.58}_{-0.84}\\
	\enddata
\end{deluxetable*}

Figures \ref{fig:cornerGaussian} and \ref{fig:cornerPowerlaw} show the one- and two-dimensional marginalized views of the posterior distribution for the Gaussian. and power law jet model fits, respectively.  Table \ref{tab:MCMC} additionally gives constraints on the total jet energy \Etot{} and the ratio $\thobs/\thC$. The parameters $E_0$, $n_0$, $\epse$, and $\epsB$ are only constrained to within an order of magnitude, due partially to the observed radio, optical, and X-ray data all lying on the same synchrotron power-law segment.  These parameters are shared between jet models, and are broadly consistent between the two fits with the possible exception of $n_0$.  The electron energy index $p$ is extremely well constrained by both models, of course, due precisely to the large range of data laying on the same synchrotron segment.  The total jet energy in both models in constrained to be on the order $10^{51}$erg.

Of most interest are the geometric parameters: the viewing angle $\thobs$ and the jet structure parameters $\thC$, $\thW$, and $b$.  Encouragingly, the fits agree with the back-of-the-envelope reasoning from the analytic closure relations and break time scalings.  Both models constrain $\thobs$ and $\thC$ reasonably well, but constrain the combination $\thobs/\thC$ far better.  This is very evident by the corresponding map of the posterior in Figures \ref{fig:cornerGaussian} and \ref{fig:cornerPowerlaw}, which clearly displays a preferred linear relationship between $\thobs$ and $\thC$.  This is the manifestation of the structured jet closure relations with $g \approx \geff(\thobs/\thC)$.  The looser constraint on $\thobs/\thC$ in the power law jet comes from the additional freedom of the $b$ parameter, which has an effect on $\geff$ in that model (see Equation \eqref{eq:geff}). The truncation angle $\thW$ is essentially unconstrained in the Gaussian fit as a far off-axis phase was not observed.  In the power law fit, however, $\thW$ is constrained to be quite narrow so as to avoid the early bright wings.

Despite the tight constraints each model gives on $\thobs$, $\thC$, and especially $\thobs/\thC$, the resulting posteriors are incompatible with each other.  The median and 68\% uncertainties in Table \ref{tab:MCMC} display only a mild tension in $\thobs$  and $\thC$ themselves, but a very large discrepancy in $\thobs/\thC$.  This is, of course, due to the very different energy profiles $E(\theta)$ in each model.  Both models can easily accommodate a rising light curve $F_\nu \sim t^{0.9}$, that is produce an effective structure parameter $\geff = 7.5$, but do so using very different geometries.

>>>>

Interpreting the afterglow as arising from a structured jet allows one to draw the following quick conclusions:
\begin{enumerate}
	\item The spectral slope $\beta = 0.585 \pm 0.005$ throughout the entire evolution, identifying the relevant spectral regime as G: $\nu_m < \nuobs < \nu_c$, and $p = 2.17 \pm 0.01$ \citep{Troja:2019ab}.
	\item The initial rising ($\alpha > 0$) light curve with declining spectrum ($\beta < 0$) immediately identify this as a \emph{misaligned} evolution ($\thobs > \thC$).  Aligned jets may only have a rising light curve when viewed at frequencies below the peak frequency, that is when $\beta < 0$.
	\item The rising light curve maintains $\alpha  = 0.90 \pm 0.06$ \citep{Troja:2019ab}.  Using the closure relations from Table \ref{tab:slopes} we can identify this as the \emph{structured} phase and determine $\geff = 8.2 \pm 0.5$.  If the light curve underwent an \emph{far-off-axis} phase, it concluded before the afterglow was detected: $\tW < 9$d.  The peak can be identified as the jet-break: $\tb = 164 \pm 12$ d.
	\item In the context of a Gaussian jet, the measurement of $\geff$ leads immediately to $\thobs / \thC = 5.7 \pm 0.2$ using Equation \eqref{eq:geff}.  A $b=2$ power law jet cannot achieve this $\geff$ for any reasonable value of $\thobs / \thC$.  A $b=6$ power law jet requires $\thobs/\thC = 14 \pm 1$, while a $b=9$ power law requires $\thobs/\thC = 9.6\pm 0.5$.
\end{enumerate}
To go further we first need to assume a value for $E_0 / n_0$. For demonstration we use the \citet{Fong:2015aa} fiducial values of $E_0 \sim 2\times 10^{51}$erg and $n_0 \sim 10^{-2}$cm$^{-3}$, assuming 1 dex uncertainty on both parameters.
\begin{enumerate}
	\setcounter{enumi}{4}
	\item We can use the jet break $\tb$ measurement to determine $\thobs + \thC$ using Equation \eqref{eq:tbout}. Using our assumed values for  $E_0$ and $n_0$ leads to $\thobs + \thC \approx 0.92 \pm 0.38$ rad ($53^\circ \pm 22^\circ$). This does not depend on the jet model.
	\item For a Gaussian jet the combined measurements of $\thobs +\thC$ and $\thobs/\thC$ then imply $\thC \approx 0.14 \pm 0.06$ rad ($8.0^\circ  \pm 3.4^\circ$) and $\thobs \approx 0.78 \pm 0.32$ rad ($45^\circ \pm 18^\circ$).
	\item For a $b=6$ power law jet we get $\thC \approx 0.063 \pm 0.026$ rad ($3.6^\circ  \pm 1.5^\circ$) and $\thobs \approx 0.86 \pm 0.35$ rad ($49^\circ \pm 20^\circ$).
\end{enumerate}

@@

 Analysis of the \gwbns{} gravitational wave signal produced estimates of the inclination angle $\iota$ between the binary neutron stars' orbital plane and our line of sight, but were dependent on the assumed value of the Hubble constant $H_0$ \citep{Abbott:2017aa}.  The values obtained were:
 $\iota = \left({28}^{+27}_{-20}\right)^\circ$ (90\% uncertainties, marginalized over $H_0$), $\iota = \left(26^{+12}_{-15}\right)^\circ$ ($H_0$ from SHoES: \citet{Riess:2016aa}), and $\iota = \left(19^{+12}_{-12}\right)^\circ$ ($H_0$ from \citet{Planck-Collaboration:2016aa}), we have reversed the reported orientation of $\iota$ to ease comparison with $\thobs$.

If the BNS merger produces a jet orthogonal to the orbital plane then $\iota = \thobs$ and we can directly compare the gravitational and electromagnetic estimates.  With our fiducial values for $E_0$ and $n_0$ the estimated value of $\thobs$ from the afterglow is larger than the gravitational wave $\iota$ measurements, but given the large uncertainties on both values there is no tension in the discrepancy.

There is much more information in the afterglow light curve than just the rising slope and jet break.  For a more detailed analysis we performed Bayesian parameter estimation utilizing \afterglowpy{} and the \emcee{} Markov chain Monte Carlo sampler \citep{Foreman-Mackey:2013aa}.  We used the light curve data reported in \citet{Troja:2019ab} with additional \hubble{} observations reported in \citet{Lamb:2019aa}. Reported measurement uncertainties were assumed to be independent and Gaussian, upper limits were treated as observations of zero flux with a 1-sigma Gaussian uncertainty. We assume $\iota = \thobs$ and use the posterior probability distribution $p_{\mathrm{LIGO}}(\cos \iota)$ reported in \citet{Abbott:2017aa} (assuming Planck $H_0$, their Figure 3) as a prior for $\thobs$ as in \citet{Troja:2018aa}.

We fit both the Gaussian and power law structured jet models, fixing $\xi_N = 1$ and $\dL = 1.23 \times 10^{26}$ cm.  The fit parameters, their priors, and their marginalized posteriors are given in Table \ref{tab:MCMC}. Both models were run with 300 walkers for 64 000 iterations, discarding the first 16 000 iterations as a burn-in.  The $E_0$, $n_0$, $\epse$, and $\epsB$ parameters were given $\log$-uniform priors, while $\thC$, $\thW$, $b$, and $p$ were given uniform priors.  The specific bounds and priors for each parameter are reported in Table \ref{tab:MCMC}.

\begin{deluxetable*}{CCCCCC}
	\tablecaption{Parameter estimation priors and marginalized posteriors for the \gwbns{} afterglow using the \afterglowpy{} Gaussian and power law jet models, including viewing angle constraints from LIGO assuming the \planck{} value of $H_0$.  Given posterior values for each model are the median, 16\%, and 84\% quantiles.  Parameters in the lower section are derived from the posterior distributions of the fit parameters in the upper sections. \label{tab:MCMC}}
	\tablehead{\colhead{Parameter}& \colhead{Unit} & \colhead{Prior Form} & \colhead{Bounds} & \colhead{Gaussian Jet Posterior} & \colhead{Power Law Jet Posterior}}
	\startdata
		\thobs	& \text{rad	} & \sin \thobs \times p_\mathrm{LIGO}(\cos \thobs)	& [0, 0.8] 	& 0.40^{+0.11}_{-0.11} 	& 0.44^{+0.12}_{-0.12} \\[0.05cm]
		\log_{10} E_0	& \text{erg	}	& \text{uniform}	& [45, 57] 						& 52.96^{+0.97}_{-0.72} 	& 52.93^{+1.1}_{-0.75}\\[0.05cm]
		\thC			& \text{rad	}	& \text{uniform}	& [0.01, $\pi/2$]					& 0.066^{+0.018}_{-0.018} & 0.046^{+0.013}_{-0.013}\\[0.05cm]
		\thW			& \text{rad}	& \text{uniform}	& [0.01, 12$\thC$]				& 0.47^{+0.26}_{-0.19} 	& 0.238^{+0.071}_{-0.69}\\[0.05cm]
		b			& - 			& \text{uniform}	& [0, 10] 						& - 					& 9.03^{+0.70}_{-1.1}\\[0.05cm]
		\log_{10}n_0	& \text{cm}^{-3} & \text{uniform}	& [-10, 10] 					& -2.70^{+0.95}_{-1.0} 	& -2.6^{+1.1}_{-1.1}\\[0.05cm]
		p			& - 			& \text{uniform}	& [2, 5] 						& 2.168^{+0.063}_{-0.0075} & 2.1653^{+0.0085}_{-0.010}\\
		\log_{10}\epse& - 			& \text{uniform}	& [-5, 0] 						& -1.42^{+0.70}_{-1.1} 	&-1.24^{+0.73}_{-1.2} \\[0.05cm]
		\log_{10}\epsB	& - 			& \text{uniform}	& [-5, 0] 						& -3.96^{+1.1}_{-0.74} 	& -3.76^{+1.1}_{-0.87}\\[0.05cm]
		\hline
		\log_{10}\Etot	& \text{erg}	&	-		&	-						& 50.57^{+0.92}_{-0.66} 	&50.46^{+1.1}_{-0.73} \\[0.05cm]
		\thobs/\thC	& -			&	-		&	-						& 6.12^{+0.18}_{-0.18} 	& 9.38^{+0.73}_{-0.56}\\[0.05cm]
		\log_{10} E_0/n_0 &  \text{erg cm}^3	&	- 	&	-						& 55.69^{+1.1}_{-0.85} 	& 55.62^{+1.2}_{-0.83}\\[0.05cm]
	\enddata
\end{deluxetable*}

Table \ref{tab:MCMC} gives the posterior median and 68\% quantiles found for each parameter, as well as constraints on the total jet energy \Etot{} and the ratio $\thobs/\thC$. Figures \ref{fig:cornerGaussian} and \ref{fig:cornerPowerlaw} show the one- and two-dimensional marginalized views of the posterior distribution for the Gaussian and power law jet model fits, respectively.  The parameters $E_0$, $n_0$, $\epse$, and $\epsB$ are only constrained to within an order of magnitude, due partially to the observed radio, optical, and X-ray data all lying on the same synchrotron power-law segment.  These parameters are shared between jet models, and are consistent between the two fits.  The electron energy index $p$ is extremely well constrained by both models, of course, due precisely to the large range of data laying on the same synchrotron segment.  The total jet energy in both models is constrained to be on the order $3 \times 10^{50}$ erg, with uncertainties of an order of magnitude.

Of most interest are the geometric parameters: the viewing angle $\thobs$ and the jet structure parameters $\thC$, $\thW$, and $b$. Both models constrain $\thobs$ and $\thC$ reasonably well, $\thobs = 0.40 \pm 0.11$ rad for the Gaussian jet and $\thobs = 0.44 \pm 0.12$ rad for the power law, but constrain the combination $\thobs/\thC$ far better: $\thobs/\thC  = 6.12 \pm 0.18$ for the Gaussian and $\thobs/\thC  = 9.38^{+0.73}_{-0.56}$ for the power law. This is very evident by the corresponding map of the posterior in Figures \ref{fig:cornerGaussian} and \ref{fig:cornerPowerlaw}, which clearly displays a preferred linear relationship between $\thobs$ and $\thC$. This is the manifestation of the structured jet closure relations with $g \approx \geff(\thobs/\thC)$.  The looser constraint on $\thobs/\thC$ in the power law jet comes from the additional freedom of the $b$ parameter, which has an effect on $\geff$ in that model (see Equation \eqref{eq:geff}). The truncation angle $\thW$ is essentially unconstrained in the Gaussian fit as a far off-axis phase was not observed.  In the power law fit, however, $\thW$ is constrained to be quite narrow so as to avoid the early bright wings.

The fits agree with the back-of-the-envelope reasoning from the analytic closure relations and jet break time.  The closest agreement is in the $\thobs/\thC$ ratio, the Gaussian agrees within the uncertainties and the power law fit agrees well with the $b=9$ estimate. The absolute values of $\thobs$ and $\thC$ agree within uncertainties but are systematically higher in the back-of-the-envelope estimates compared to the fit results.  This can be accounted for by our adopted fiducial value for $E_0/n_0$ which in the estimate was taken to be $10^{53.3}$ erg cm$^3$.  The MCMC fits, which constrain this value for this particular burst, both find $E_0/n_0 \approx 10^{55.6 \pm 1.0}$ erg cm$^{3}$.  Adopting this value in the back-of-the-envelope estimates leads to $\thobs = 0.41\pm0.12$ rad, $\thC = 0.07 \pm 0.02$ rad for the Gaussian and $\thobs = 0.43\pm0.12$ rad, $\thC = 0.05 \pm 0.01$ for the power law in excellent agreement with the MCMC fits.

% TODO: RECONSIDER, CLARIFY, OR DROP THIS
Despite the tight constraints each model gives on $\thobs$, $\thC$, and especially $\thobs/\thC$, the resulting posteriors are incompatible with each other.  The median and 68\% uncertainties in Table \ref{tab:MCMC} display only a mild tension in $\thobs$  and $\thC$ themselves, but a very large discrepancy in $\thobs/\thC$.  This is, of course, due to the very different energy profiles $E(\theta)$ in each model.  Both models can easily accommodate a rising light curve $F_\nu \sim t^{0.9}$, that is produce an effective structure parameter $\geff = 7.5$, but do so using very different geometries.
